---
title: "Pitcher X Analysis"
author: "Jack Weyer"
date: "4/12/2022"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---
```{r}
library(tidyverse)
df <- read.csv("C:/Users/jack_/Downloads/Pitcher X Data 2022.csv")
df = df %>%
  mutate(Date = as.Date(Date, format='%m/%d/%Y'))
```

```{r}
stats <- function(x){
  x %>%
  summarise(`SwStr%` = sum(whiff_prob)/n(), n=n(), swingProb = mean(swing_prob), `Whiff%`=sum(whiff_prob)/sum(swing_prob))
}
# SwStr% is % of pitches that are whiffs... what we are trying to optimize.. 0.87 correlation with K% (insiderbaseball.com)
# Whiff% if % of swings that miss
sum(df$whiff_prob)/sum(df$swing_prob) #RIGHT!!!!!!!!!!!!!!!!!!11
sum(df$whiff_prob/df$swing_prob) #WRONG!!!!!!!!
df %>%
  stats()
  
totalxwhiffs = 0.1155503*2716
#314 total xWhiffs


# 11.6 SwStr%... 21st out of 40 qualified pitchers (FanGraphs)
```
```{r}
msBlue = '#0C2C56'
msGreen = '#005C5C'
df %>%
  group_by(Date) %>%
  summarise(n=n())
```


```{r}
whiff <- read_csv("C:/Users/jack_/Downloads/stats.csv")
whiff %>%
  ggplot(aes(x = reorder(last_name, -whiff_percent), y=whiff_percent, 
             fill=factor(ifelse(last_name=='X', "Highlighted","Normal"))))+
  geom_bar(stat='identity')+
  theme_minimal()+
  scale_fill_manual(name="last_name", values=c(msGreen,msBlue))+
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        legend.position = 'none',
        axis.title.y = element_text(size=14))+
  scale_y_continuous(labels=scales::percent_format(scale=1))+
  labs(x='', y='xWhiff% (xWhiffs per Swing)',
       title = 'Pitcher X was NOT elite when the batter swings (2021 qualifiers)')
```

```{r}
swStr <- read_csv("C:/Users/jack_/Downloads/FanGraphs Leaderboard.csv")
swStr %>%
  ggplot(aes(x = reorder(Name, -`SwStr%`), y=`SwStr%`, 
             fill=factor(ifelse(Name=='Pitcher X', "Highlighted","Normal"))))+
  geom_bar(stat='identity')+
  theme_minimal()+
  scale_fill_manual(name="Name", values=c(msGreen,msBlue))+
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        legend.position = 'none',
        axis.title.y = element_text(size=14))+
  labs(x='', y='xSwStr%',
       title = 'Pitcher X was below average on whiffs overall (2021 qualifiers)')
```



```{r}
makedate <- function(x){
  return(as.Date(x, format='%m/%d/%Y'))
}
```


```{r}
dateRates = df %>%
  group_by(Date) %>%
  stats()

Allstar = as.Date('07/13/2021', format='%m/%d/%Y')


dateRates %>%
  ggplot(aes(x=Date, y=`SwStr%`))+
  geom_point()+
  geom_line()+
  #geom_vline(xintercept = Allstar, color='red', size=2)+
  geom_segment(aes(x=makedate('04/04/2021'), xend=Allstar, y=0.1209653, yend=0.1209653), color= msBlue, size=2)+
  geom_segment(aes(x=Allstar, xend=makedate('10/03/2021'), y=0.110331, yend=0.110331), color= msGreen, size=2)+
  scale_y_continuous(labels=scales::percent_format(scale=100))+
  labs(title="Pitcher X's whiffs per pitch dropped off after the All-Star break",
       y='xSwStr%')+
  theme_minimal()
```
No noticable trend. Will group by before/after all star game
```{r}
df %>%
  filter(Date > Allstar) %>%
  summarise(swstr = sum(whiff_prob)/n())
```

#########################################
```{r}
df %>%
  group_by(Inning) %>%
  summarise(xSwStr = sum(whiff_prob)/n()) %>%
  ggplot(aes(x=Inning, y=xSwStr))+
  geom_point()+
  geom_line()
```
No inning trend
```{r}
df %>%
  group_by(PAofInning) %>%
  filter(PAofInning<9) %>%
  summarise(xWhiffRate = n()/sum(whiff_prob)) %>%
  ggplot(aes(x=PAofInning, y=xWhiffRate))+
  geom_point()+
  geom_line()
```
Can't say anything from this. Correlated with opposing team batting order

```{r}
df %>%
 group_by(BatterSide) %>%
  stats()
```
There is a platoon difference. Better against righties

```{r}
df %>%
 group_by(PitchofPA) %>%
  summarise(xWhiffRate = n()/sum(whiff_prob)) %>%
  ggplot(aes(x=PitchofPA, y=xWhiffRate))+
  geom_point()+
  geom_line()
```
TONS of first pitch takes. Should actually consider removing first pitches from df for more actionable insights

```{r}
countWhiffs = df %>%
 group_by(Balls, Strikes) %>%
  stats()
```
```{r}
library(scales)
countWhiffs %>%
  ggplot(aes(x=Balls, y=Strikes, fill=`SwStr%`))+
  geom_tile()+
  geom_text(aes(label=percent(`SwStr%`, accuracy=1)), color='black')+
  scale_fill_distiller(palette = 'RdBu',direction=-1)+
  guides(fill=guide_colourbar(barwidth=.5, barheight = 15))+
  labs(title="Pitcher X's xSwStr% (Whiffs per Pitch) by Count")+
  theme_classic()
```

```{r}
df %>%
  group_by(Balls, Strikes) %>%
  summarise(n=n()) %>%
  mutate(rate=n/716) %>%
  ggplot(aes(x=Balls, y=Strikes, fill=rate))+
  geom_tile()+
  geom_text(aes(label=percent(`rate`, accuracy=1)), color='black')+
  scale_fill_distiller(palette = 'RdBu',direction=-1)+
  guides(fill=guide_colourbar(barwidth=.5, barheight = 15))+
  labs(title="% of 2021 Plate Appearances that reach given count")+
  theme_classic()
```

```{r}
df %>%
  group_by(Date, Inning, PAofInning) %>%
  summarise(n=n())
```
716 batters faced




Pitcher's count
```{r}
pitcherCount = df[(df$Balls==0 & df$Strikes==1) | 
                    (df$Balls==0 & df$Strikes==2) | 
                    (df$Balls==1 & df$Strikes==2) |
                    (df$Balls==2 & df$Strikes==2),]

pitcherCount %>%
  stats()
```

```{r}
hitterCount = df[(df$Balls==1 & df$Strikes==0) | 
                    (df$Balls==2 & df$Strikes==0) | 
                    (df$Balls==2 & df$Strikes==1) |
                    (df$Balls==3 & df$Strikes==0) |
                    (df$Balls==3 & df$Strikes==1),]

hitterCount %>%
  stats()
```


```{r}
countWhiffs %>%
  ggplot(aes(x=Balls, y=Strikes, fill=`Whiff%`))+
  geom_tile()+
  geom_text(aes(label=percent(`Whiff%`, accuracy=1)), color='black')+
  scale_fill_distiller(palette = 'RdBu',direction=-1)+
  guides(fill=guide_colourbar(barwidth=.5, barheight = 15))+
  labs(title="Pitcher X's xWhiff% (Whiffs per Swing) by Count")+
  theme_classic()
```


No whiffs on 0 strike counts (same as above)
```{r}
swingProbs = df %>%
  group_by(Balls, Strikes) %>%
  summarise(Prob = mean(swing_prob))
```
```{r}
swingProbs %>%
  ggplot(aes(x=Balls, y=Strikes, fill=Prob))+
  geom_tile()+
  geom_text(aes(label=percent(Prob,2)), color='black')+
  scale_fill_viridis_c(option='I')+
  guides(fill=guide_colourbar(barwidth=.5, barheight = 15))+
  labs(title="Pitcher X's Swing Probability by Count (2021)")+
  theme_classic()
```

```{r}
countWhiffs %>%
  ggplot(aes(x=swingProb, y=`SwStr%`))+
  geom_point()+
  geom_smooth(method='lm')+
  labs(title='Higher swing% is associated with more whiffs per pitch')
```



Swing probability is at its lowest by far on 0-0 counts (besides 3-0 of course)
```{r}
df %>%
  group_by(Balls, Strikes) %>%
  summarise(n=n())
```
PAs ended on his first pitch only 13% of the time! 
59% of non-play first pitch's went for strikes... this needs to increase.

```{r}
df %>%
  filter(Balls==3, Strikes==2) %>%
  group_by(PitchType) %>%
  summarise(n=n())
```
```{r}
df %>%
  filter(Balls==0, Strikes==0) %>%
  ggplot(aes(x=PlateSide, y=PlateHeight, color=swing_prob))+
  coord_fixed(ratio=1)+
  geom_point(alpha=.8)+
  scale_color_distiller(palette = 'YlOrRd',direction=1)+
  guides(color=guide_colourbar(barwidth=.5, barheight = 15))+
  geom_rect(xmin=(-17/2/12), xmax=17/2/12, ymin=1.5, ymax=3.5, alpha=.005)+
  labs(title='0-0 outcomes: 51% strike, 36% ball, 13% in play', 
       color = 'Swing Probability',
       x='feet',
       y='feet')+
  theme_dark()
```
# Threw 16/20 3-0 pitches in the zone (75%).. capable of throwing in the zone when needed


```{r}
arsenalByPitch = df %>%
  group_by(Balls, Strikes, PitchType) %>%
  summarise(n=n()) %>%
  mutate(rate = n/sum(n))
```
```{r}
df %>%
  group_by(Balls, Strikes, PitchType) %>%
  stats()
```





```{r}
df %>%
  filter(Balls==0, Strikes==0) %>%
 group_by(PitchType) %>%
  summarise(xWhiffRate = n()/sum(whiff_prob), n=n())
```


STRATEGY... 
so many takes on first pitch..
1-0 count (8.8 whiff rate)
0-1 count (7.3 whiff rate)
therefore...
attack the zone first pitch, get called strikes to generate 0-1 counts and thus more whiffs!

Whiff rates are generally higher with more strikes in count


#####################Strategy 2... less fastballs############

```{r}
df %>%
 group_by(PitchType) %>%
  stats()
```
Arsenal..
Fastball  40%
Cutter    29%
Changeup  16%
Curveball 15% 

Want more whiffs? Throw more changeups, cutters and less fastballs, curveballs

```{r}
hittables = df[!(df$Balls==3 & df$Strikes==0),]

hittables = hittables %>%
  filter(PitchofPA != 1) %>%
  group_by(PitchType) %>%
  stats()

hittables %>%
  ggplot(aes(x=n, y=reorder(PitchType, n), fill=`SwStr%`))+
  scale_fill_distiller(palette = 'RdBu',direction=-1)+
  geom_bar(stat='identity')+
  labs(title='Non 0-0 and 3-0 counts',
       x = 'Total Pitches (2021)',
       y='',
       fill='xSwStr%')+
  geom_text(aes(label=percent(round(`SwStr%`,3))), hjust=1.3)+
  theme_classic()
```
In non-super low swing probability counts, whiff rate is still worst with X's fastball. His changeup is best here

FB 36%
Cutter 33%
Changeup 18%
Curveball 13%

Curveball.... great 3-2 pitch
```{r}
ThreeTwo = df %>%
  filter(Balls==3, Strikes==2) %>%
  group_by(PitchType) %>%
  stats()

ThreeTwo %>%
  mutate(expextation = sum(n*`SwStr%`))
```
3-2 statustics
Fastball  51%
Cutter    36%
Changeup  10%
Curveball 3%

New-strat
```{r}
48*0.09622405 + 42*0.15328203 + 35*0.19529297 + 14*0.16335887
```
Two more xWhiffs on the season


Only went to it 4 times!!! Huge inefficiency there
```{r}
hittables %>%
  filter(PitchofPA != 1) %>%
  group_by(PitchType, BatterSide) %>%
  summarise(xWhiffRate = n()/sum(whiff_prob), n=n(), swingProb = mean(swing_prob), WhiffgivenSwing=mean(whiff_prob_gs))
```


Case study
```{r}
df %>%
  filter(Balls==3, Strikes==1) %>%
  group_by(PitchType) %>%
  stats()
```

```{r}
df %>%
  filter(Balls==1, Strikes==2) %>%
  group_by(PitchType) %>%
  stats()
```

```{r}
df %>%
  group_by(PitchType) %>%
  summarise(meanspeed = mean(ReleaseSpeed))
```

